---
title: "RNA-seq Differential Expression Report"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 3
    toc-float: true
    theme: flatly
    code-fold: true
    self-contained: true
  pdf:
    toc: true
    toc-depth: 3
    geometry: margin=1in
    fontsize: 11pt
    colorlinks: true
params:
  results: NULL
  significant: NULL
  contrast_level: "contrast"
  ref_level: "reference"
  padj_cutoff: 0.05
  lfc_cutoff: 1
  go_results: NULL
  kegg_results: NULL
  llm_interpretation: NULL
  llm_upregulated: NULL
  llm_downregulated: NULL
  model_used: "mistral"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 9,
  fig.height = 6
)

library(dplyr)
library(ggplot2)
library(ggrepel)
library(knitr)
library(kableExtra)

# Plain objects from params — no S4
res            <- params$results
sig            <- params$significant
go_res         <- params$go_results
kg_res         <- params$kegg_results
contrast_level <- params$contrast_level
ref_level      <- params$ref_level
padj_cutoff    <- params$padj_cutoff
lfc_cutoff     <- params$lfc_cutoff
```

# Experiment Overview

```{r summary-table}
data.frame(
  Parameter = c(
    "Contrast",
    "Reference level",
    "Total genes tested",
    "Significant DE genes",
    "Upregulated",
    "Downregulated",
    "Adj. p-value cutoff",
    "Log2FC cutoff"
  ),
  Value = c(
    contrast_level,
    ref_level,
    nrow(res),
    nrow(sig),
    sum(sig$log2FoldChange > 0),
    sum(sig$log2FoldChange < 0),
    padj_cutoff,
    lfc_cutoff
  )
) |>
  kable(align = c("l", "r")) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width        = FALSE,
    position          = "left"
  )
```

---

# Differential Expression Results

## Volcano Plot

```{r volcano-plot}
res_plot <- res |>
  mutate(
    neg_log10_padj = -log10(padj),
    status = case_when(
      padj < padj_cutoff & log2FoldChange >  lfc_cutoff ~ "Up",
      padj < padj_cutoff & log2FoldChange < -lfc_cutoff ~ "Down",
      TRUE ~ "NS"
    )
  )

top_genes <- res_plot |>
  filter(status != "NS") |>
  arrange(padj) |>
  slice_head(n = 15)

ggplot(res_plot, aes(log2FoldChange, neg_log10_padj, color = status)) +
  geom_point(alpha = 0.6, size = 1.2) +
  geom_text_repel(
    data = top_genes,
    aes(label = gene),
    size = 3,
    max.overlaps = 20
  ) +
  geom_vline(xintercept = c(-lfc_cutoff, lfc_cutoff),
             linetype = "dashed", color = "grey50") +
  geom_hline(yintercept = -log10(padj_cutoff),
             linetype = "dashed", color = "grey50") +
  scale_color_manual(
    values = c(Up = "#E64B35", Down = "#4DBBD5", NS = "grey70")
  ) +
  labs(
    x     = "Log2 Fold Change",
    y     = "-Log10 Adjusted P-value",
    color = "Status",
    title = paste("Volcano Plot:", contrast_level, "vs", ref_level)
  ) +
  theme_bw(base_size = 13) +
  theme(legend.position = "top")
```

## Top Upregulated Genes

```{r up-table}
sig |>
  filter(log2FoldChange > 0) |>
  arrange(padj) |>
  slice_head(n = 20) |>
  mutate(
    log2FoldChange = round(log2FoldChange, 3),
    baseMean       = round(baseMean, 1),
    padj           = signif(padj, 3),
    pvalue         = signif(pvalue, 3)
  ) |>
  dplyr::select(gene, baseMean, log2FoldChange, pvalue, padj) |>
  kable(col.names = c("Gene", "Base Mean", "Log2FC", "P-value", "Adj. P-value")) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width        = FALSE
  ) |>
  row_spec(0, bold = TRUE)
```

## Top Downregulated Genes

```{r down-table}
sig |>
  filter(log2FoldChange < 0) |>
  arrange(padj) |>
  slice_head(n = 20) |>
  mutate(
    log2FoldChange = round(log2FoldChange, 3),
    baseMean       = round(baseMean, 1),
    padj           = signif(padj, 3),
    pvalue         = signif(pvalue, 3)
  ) |>
  dplyr::select(gene, baseMean, log2FoldChange, pvalue, padj) |>
  kable(col.names = c("Gene", "Base Mean", "Log2FC", "P-value", "Adj. P-value")) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width        = FALSE
  ) |>
  row_spec(0, bold = TRUE)
```

## Full Results Table

```{r full-table}
res |>
  arrange(padj) |>
  mutate(
    log2FoldChange = round(log2FoldChange, 3),
    baseMean       = round(baseMean, 1),
    padj           = signif(padj, 3),
    pvalue         = signif(pvalue, 3)
  ) |>
  dplyr::select(gene, baseMean, log2FoldChange, pvalue, padj) |>
  kable(col.names = c("Gene", "Base Mean", "Log2FC", "P-value", "Adj. P-value")) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width        = TRUE
  ) |>
  scroll_box(height = "400px")
```

---

# Pathway Enrichment

## GO Biological Process

```{r go-plot, fig.height=8}
if (!is.null(go_res) && nrow(go_res) > 0) {
  go_top <- go_res |>
    filter(p.adjust < 0.05) |>
    arrange(p.adjust) |>
    slice_head(n = 20) |>
    mutate(
      Description = factor(Description, levels = rev(Description)),
      GeneRatio_num = sapply(GeneRatio, function(x) {
        parts <- strsplit(x, "/")[[1]]
        as.numeric(parts[1]) / as.numeric(parts[2])
      })
    )

  ggplot(go_top, aes(GeneRatio_num, Description, color = p.adjust, size = Count)) +
    geom_point() +
    scale_color_gradient(low = "#E64B35", high = "#4DBBD5") +
    labs(
      x     = "Gene Ratio",
      y     = NULL,
      color = "Adj. P-value",
      size  = "Gene Count",
      title = "GO Biological Process Enrichment"
    ) +
    theme_bw(base_size = 11) +
    theme(axis.text.y = element_text(size = 9))
} else {
  cat("No significant GO terms detected.")
}
```

## KEGG Pathways

```{r kegg-plot, fig.height=8}
if (!is.null(kg_res) && nrow(kg_res) > 0) {
  kg_top <- kg_res |>
    filter(p.adjust < 0.05) |>
    arrange(p.adjust) |>
    slice_head(n = 20) |>
    mutate(
      Description = factor(Description, levels = rev(Description)),
      GeneRatio_num = sapply(GeneRatio, function(x) {
        parts <- strsplit(x, "/")[[1]]
        as.numeric(parts[1]) / as.numeric(parts[2])
      })
    )

  ggplot(kg_top, aes(GeneRatio_num, Description, color = p.adjust, size = Count)) +
    geom_point() +
    scale_color_gradient(low = "#E64B35", high = "#4DBBD5") +
    labs(
      x     = "Gene Ratio",
      y     = NULL,
      color = "Adj. P-value",
      size  = "Gene Count",
      title = "KEGG Pathway Enrichment"
    ) +
    theme_bw(base_size = 11) +
    theme(axis.text.y = element_text(size = 9))
} else {
  cat("No significant KEGG pathways detected.")
}
```

---

# LLM-Assisted Biological Interpretation

> **Note:** The following sections were generated by a locally deployed large language
> model (`r params$model_used`) based on the differential expression and pathway
> enrichment results above. All interpretations should be critically evaluated
> in the context of your experimental system.

---

## Overall Biological Interpretation

```{r llm-overall, results='asis'}
if (!is.null(params$llm_interpretation)) {
  cat(params$llm_interpretation)
} else {
  cat("LLM interpretation not available.")
}
```

---

## Upregulated Genes — Biological Context

```{r llm-up, results='asis'}
if (!is.null(params$llm_upregulated)) {
  cat(params$llm_upregulated)
} else {
  cat("LLM interpretation not available.")
}
```

---

## Downregulated Genes — Biological Context

```{r llm-down, results='asis'}
if (!is.null(params$llm_downregulated)) {
  cat(params$llm_downregulated)
} else {
  cat("LLM interpretation not available.")
}
```

---

# Session Information

```{r session-info}
sessionInfo()
```
